{\smaller\setlength{\parskip}{0.8em}\setlength{\parindent}{0em}%
{\raggedright%

\bookTitle \\
by \bookAuthor

Published by:

Aruna Publications,\\
Aruna Ratanagiri Buddhist Monastery,\\
2 Harnham Hall Cottages,\\
Harnham, Belsay,\\
Northumberland, NE20 0HF\\
UK

Contact Aruna Publications at \href{http://aruno.org}{www.aruno.org}\\
This book is available for free download at\\
\href{http://forestsanghapublications.org/}{www.forestsanghapublications.org}

ISBN \bookISBN

Copyright \copyright\ 2011 HARNHAM BUDDHIST MONASTERY TRUST

\vfill

{\tiny

This work is licenced under the Creative Commons Attribution-NonCommercial-NoDerivs 2.0 UK: England \& Wales Licence. To view a copy of this licence, visit:\\
\href{http://creativecommons.org/licenses/by-nc-nd/2.0/uk/}{http://creativecommons.org/licenses/by-nc-nd/2.0/uk/}\\
Or send a letter to: Creative Commons, 444 Castro Street, Suite 900, Mountain View, California, 94041, USA.

See page \pageref{cc-details} for more details on your rights and restrictions under this licence.

Material included in this book has been previously published by Wat Pah Nanachat, Thailand, reprinted here with permission.

Copyright \copyright\ WAT PAH NANACHAT

ISBN 978-616-90279-6-6

Cover photo taken at Wat Dtao Dam, a branch monastery of Wat Pah Nanachat near the boarder between Thailand and Burma.

Leaf drawing by Madeleine Gusdal.

Produced with the {\fontfamily{cms}\selectfont\LaTeX} typesetting system. The body-text is typeset in Gentium, distributed with the SIL Open Font Licence by SIL International.

\editioninfo

Printed in Malaysia by Bolden Trade.

}

}}

