
Meditate reciting `\textit{Buddho, Buddho}'\footnote{The word \textit{Buddho} is often taught by meditation masters of the Thai forest tradition as a word to recite mentally in combination with the breath. One recites the syllable `Bud' on the in-breath and `dho' on the out-breath.} until the word penetrates deep into the heart of your consciousness (\textit{citta}). \textit{Buddho} represents the awareness and wisdom of the Buddha. When practising, you must depend on this word more than anything else. The awareness it brings will lead you to understand the truth about your own mind. It is a true refuge, which means that both mindfulness and insight are present.

Even wild animals can have awareness of a sort. They have mindfulness as they stalk their prey and prepare to attack. The predator needs firm mindfulness to keep hold of the captured prey, however defiantly it struggles to escape death. That is one kind of mindfulness. For this reason you must be able to distinguish between different kinds of mindfulness. \textit{Buddho} is a way to apply the mind. When you consciously apply the mind to an object, it wakes up. The awareness wakes it up. Once this knowing has arisen through meditation, you can see the mind clearly. As long as the mind remains without the awareness of \textit{Buddho}, even if ordinary worldly mindfulness is present, it is as if unawakened and without insight. It will not lead you to what is truly beneficial.

Mindfulness depends on the presence of \textit{Buddho} -- the knowing. It must be a clear knowing, which leads to the mind becoming brighter and more radiant. The illuminating effect that this clear knowing has on the mind is similar to the brightening of a light in a darkened room. As long as the room is pitch-black, any objects placed inside remain difficult to distinguish or are completely obscured from view because of the lack of light. But as you begin intensifying the brightness of the light inside, it will penetrate throughout the room, enabling you to see more clearly from moment to moment and thus allowing you to know more and more the details of any object inside.
